Processing genome scale tabular data with wormtable

### 2013-08-21 docs/paper/paper.tex
Updated abstract and made another editing pass.

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:} 
Biological research, including genomics generates huge amounts of data.
The data produced by 
programs are commonly stored as tables in plain text files and are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is an error prone and time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. 
\paragraph*{Results:} 
We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables. Files written in existing 
formats can be easily converted to wormtable format, and we 
provide conversion utilities for the VCF and GTF format.

\paragraph*{Conclusions:} Using wormtable's simple API users 
can process tables in a very high performance manner, allowing them
much greater freedom to explore their data. Our comparisons with some 
alternative software demonstrates a 100-fold increase in speed. Additionally, 
because it is no longer necessary to parse complex files,
data analysis code is also much simpler and less error prone. 
\end{abstract}

@@

It is not sufficient, however, to simply store and retrieve data. To use data
it must be \emph{processed}. This is the major flaw in 
using text files as a data format: before we can perform calculations, we 
must first parse the encoded information into native 
machine values. This is a computationally expensive process, and 
compression (if it is used) adds substantial overhead. 
As a result,  simple calculations over a large dataset may take 
many hours to complete.

Another problem with tables stored as text files is that it is difficult 
to index the information in a particular columns. 
This means that many operations on a table require a complete scan 
through the file. The only viable means of 
working with a subset of the data, therefore, 
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

The obvious solution to these problems is to load tabular data into a
relational database. Databases store values in binary form so that parsing is 
not required, and support efficient retrieval and indexing. However, there 
are many problems with this approach. Relational databases are 
complex systems, each supporting different features and SQL 
dialects. It is not a straightforward task to design a schema for a 
particular data set, particularly not if portability across different 
databases is required. Similarly, accessing data requires a knowledge
of SQL. In the most common case, a database server must also be maintained
and so storage and user permissions must be carefully managed. All of 
these aspects require significant expertise which is why so many researchers 
and programs continue to use tabular data formats.

A relational database server is far more than we require in the 
majority of cases. Data files are usually written once and 
not subsequently expected to change.
Thus, storing this information 
in a relational database with its sophisticated concurrency control
is entirely unnecessary. Centralised storage of data sets
creates an unnecessary administration burden, as does an extra
layer of user management. Therefore, although relational databases
provide powerful data management technologies, they are complex
to use and maintain and are therefore rarely used for 
static datasets.

@@

% Wormtable's binary row format 
Data is stored in rows using a compact binary format.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Each column is allocated a fixed number of bytes within the `fixed region' of 
a row. For fixed
length columns, the encoded elements are stored entirely within the
fixed region; for variable length columns, the encoded elements are 
stored in the `variable region' of the row, and the location and number 
of elements stored in the fixed region. The type, size and number 
of elements of 
each column are specified in the table schema, which is defined 
in XML as part of the table's metadata.

@@

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a 
Berkeley DB database in 
which the keys are the elements from the columns in question concatenated
together, and the values are pointers to the original row. Indexes make
many operations much more efficient, because we can go directly to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns. Moreover, wormtable supports binned indexes, in which a range of 
values is mapped to a single index key. This is very useful for floating 
point columns, where it may not be necessary to distinguish between
very similar values.

@@

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
In Python, the \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function can take two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
the rows of interest. This allows us to efficiently seek to an 
arbitrary location in the table and to read a given number 
of rows sequentially from this point.

The \texttt{Index} class also has a \texttt{cursor} function
with the same signature, but in this case, rows 
are returned in the order defined by the index. The  
\texttt{start} and \texttt{stop} arguments are now defined 
in terms of index keys: all rows in which the index key
is greater than or equal to \texttt{start} and less 
than \texttt{stop} are returned. Partial keys may also be 
provided for multi-column indexes. The \texttt{Index} 
class also provides an iterator over all keys, as well 
as a means of counting the number of rows with a given 
key.

@@

\subsection*{Scan performance}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}. The VCF format is one of 
the most commonly used formats to store genomic data from next generation sequencing.
To illustrate the advantages of wormtable when performing calculations over 
a whole table, we converted a large 
publicly available VCF file (produced as part of the
Drosophila genetic reference panel~\cite{mr+12}) to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a workstation with an Intel Xeon processor, 12GB of RAM 
and a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).

@@

When using a cold cache (i.e., no pages of the files in question are 
present in the operating system's cache) the elapsed
time for Tabix was 103 seconds with a processor time of 65
seconds. For wormtable, the elapsed time was 181 seconds 
with a processor time of 5 seconds. Thus, the time required 
to perform this test is dominated by waiting for I/O in both 
cases. Since the wormtable file is considerably larger than the 
compressed file used by Tabix, more and larger hard drive
seeks were required 
to bring the required pages into memory (solid state 
storage does not suffer from this problem). Once the 
pages were in memory, however, Tabix needed to do much more work to 
decompress then and make them usable, as shown by the 
difference in processor times.

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevant pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, the use of compression is a trade-off: 
it reduces file size, which reduces the number of random seeks 
required but decompression is expensive and must 
be repeated each time a block is accessed.

The test used here is also highly synthetic, and unlikely to be 
indicative of most real-world applications. The majority of 
workloads have strong locality of reference~\cite{d05}, 
and such large and extreme seeks across genomic regions 
are unlikely to occur. In this case, wormtable 
is  much faster than Tabix, since there
is no CPU overhead of decompression. Furthermore, since 
Tabix returns rows of text, the problem of parsing rows
must still be solved. This is expensive (as illustrated 
in the previous subsection), and creates extra code 
complexity. Wormtable, by providing a simple API to 
access both rows and columns, gives a 
unified interface for accessing data that is 
both straightforward to use and highly efficient.

@@

Wormtable is currently limited to supporting Python, and 
another important aspect of future development is to 
develop a C library along with bindings for other popular
languages such as Perl, R and PHP. 
Wormtable does not support interval search, and so it is not
straightforward to find, for example, all rows overlapping 
a given genomic region in GTF files. 
This problem has been 
solved several times, however, and we aim to adapt existing 
techniques~\cite{li11,kzbhk10} and incorporate them into 
wormtable.  Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.

>>>>

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:} 
Genomics generates huge amounts of tabular data which is 
usually stored in plain text files.
%Biological research, including genomics generates huge amounts of data.
%The data produced by 
%programs are commonly stored as tables in plain text files and are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is an error prone and time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. 
\paragraph*{Results:} 
We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables. Files written in existing 
formats can be easily converted to wormtable format, and we 
provide conversion utilities for the VCF and GTF formats.

\paragraph*{Conclusions:} 
Wormtable's simple API allows users to process large 
tables orders of magnitude more quickly than is possible 
when parsing text. Furthermore, the indexing facilities 
provide efficient access to subsets of the data along with 
providing useful methods of summarising columns. Since 
third-party libraries or custom code are no longer needed 
to parse complex plain text formats, analysis code 
can also be substantially simpler as well as being 
uniform across different data formats. These benefits of 
reduced code complexity and greatly increased performance
allow users much greater freedom to explore their data.
\end{abstract}

@@

It is not sufficient, however, to simply store and retrieve data. To use data
it must be \emph{processed}. This is the major flaw in 
using text files as a data format; before we can perform calculations, we 
must first parse the encoded information into native 
machine values. This is a computationally expensive process, and 
compression (if it is used) adds substantial overhead. 
As a result,  simple calculations over a large dataset may take 
many hours to complete.

Another problem with tables stored as text files is that it is difficult 
to index the information in particular columns. 
This means that many operations on a table require a complete scan 
through the file. The only viable means of 
working with a subset of the data, therefore, 
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

The obvious solution to these problems is to load tabular data into a
relational database. Databases store values in binary form so that parsing is 
not required, and support efficient retrieval and indexing. However, there 
are many problems with this approach. Relational databases are 
complex systems, each supporting different features and SQL 
dialects. It is not a straightforward task to design a schema for a 
particular dataset, particularly not if portability across different 
databases is required. Similarly, accessing data requires a knowledge
of SQL. In the most common case, a database server must be maintained
and so storage and user permissions must be carefully managed. All of 
these aspects require significant expertise which is why so many researchers 
and programs continue to use textual data formats.

A relational database server is far more than we require in the 
majority of cases. Data files are usually written once and 
not subsequently expected to change.
Thus, storing this information 
in a relational database with its sophisticated concurrency control
is entirely unnecessary. Centralised storage of datasets
creates an unnecessary administration burden, as does an extra
layer of user management. Therefore, although relational databases
provide powerful data management technologies, they are complex
to use and maintain and are rarely used for 
static datasets.

@@

% Wormtable's binary row format 
Data is stored in rows using a compact binary format.
Each row consists of a number of columns, and each column 
holds elements of a specific type and size.
Each column is allocated a fixed number of bytes within the `fixed region' of 
a row. For fixed
length columns, the encoded elements are stored entirely within this
fixed region; for variable length columns, the encoded elements are 
stored in the `variable region' of the row, and the location and number 
of elements stored in the fixed region. The type, size and number 
of elements of 
each column are specified in the table schema, which is defined 
in XML as part of the table's metadata.

@@

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a 
Berkeley DB database in 
which the keys are the elements from the columns in question concatenated
together, and the values are pointers to the original rows. Indexes make
many operations much more efficient, because we can go directly to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns. Moreover, wormtable supports binned indexes, in which a range of 
values is mapped to a single index key. This is very useful for floating 
point columns, where it may not be necessary to distinguish between
very similar values.

@@

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
In Python, the \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} method can take two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
the rows of interest. This allows us to efficiently seek to an 
arbitrary location in the table and to read a given number 
of rows sequentially from this point.

The \texttt{Index} class also has a \texttt{cursor} method 
with the same signature, but in this case, rows 
are returned in the order defined by the index. The  
\texttt{start} and \texttt{stop} arguments are now defined 
in terms of index keys: all rows in which the index key
is greater than or equal to \texttt{start} and less 
than \texttt{stop} are returned. Partial keys may also be 
provided for multi-column indexes. The \texttt{Index} 
class also provides an iterator over all keys, as well 
as a means of counting the number of rows with a given 
key.

@@

\subsection*{Scan performance}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}. The VCF format is one of 
the most commonly used formats to store genomic data from next generation sequencing.
To illustrate the advantages of wormtable when performing calculations over 
a whole table, we converted a large 
publicly available VCF file (produced as part of the
Drosophila genetic reference panel~\cite{mr+12}) to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program,
the VCF was converted to wormtable format 
on a workstation with an Intel Xeon processor, 12GB of RAM 
and a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).

@@

When using a cold cache (i.e., no pages of the files in question are 
present in the operating system's cache) the elapsed
time for Tabix was 103 seconds with a processor time of 65
seconds. For wormtable, the elapsed time was 181 seconds 
with a processor time of 5 seconds. Thus, the time required 
to perform this test is dominated by waiting for I/O in both 
cases. Since the wormtable file is considerably larger than the 
compressed file used by Tabix, more and larger hard drive
seeks were required 
to bring the required pages into memory. Once the 
pages were in memory, however, Tabix needed to do much more work to 
decompress then and make them usable, as shown by the 
difference in processor times.

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevant pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, the use of compression is a trade-off: 
it reduces file size, which reduces the number of random seeks 
required, but decompression is expensive and must 
be repeated each time a block is accessed.
The advantages of a smaller file in terms of reducing the
number of seeks incurred would also be largely negated 
by using solid state storage, where random seeks do not
incur such a heavy penalty.

The test used here is also highly synthetic, and unlikely to be 
indicative of most real-world applications. The majority of 
workloads have strong locality of reference~\cite{d05}, 
and such large and extreme jumps across genomic regions 
are unlikely to occur. In this case, wormtable 
is  much faster than Tabix, since there
is no CPU overhead of decompression. Furthermore, since 
Tabix returns rows of text, the problem of parsing rows
must still be solved. This is expensive (as illustrated 
in the previous subsection), and creates extra code 
complexity. Wormtable, by providing a simple API to 
access both rows and columns, gives a 
unified interface for accessing data that is 
both straightforward to use and highly efficient.

@@

Wormtable is currently limited to supporting Python, and 
another important aspect of future development is to 
create a C library along with bindings for other popular
languages such as Perl, R and PHP. 
Wormtable does not support interval search, and so it is not
straightforward to find, for example, all rows overlapping 
a given genomic region in GTF files. 
This problem has been 
solved several times, however, and we aim to adapt existing 
techniques~\cite{li11,kzbhk10} and incorporate them into 
wormtable.  Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.
